---
title: "Bayesian Bootstrap in JAGS"
output: 
  html_document:
    theme: flatly
    highlight: monochrome
---

```{r setup, include=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(boot)
library(gtools)
library(htmlwidgets)
library(metricsgraphics)
library(magrittr)
library(bayesboot)
library(rjags)
```


## Introduction

Classical bootstrap is compared with Bayesian bootstrap. The underlying difference lies in the distribution over individual sample weights. While classical bootstrap uses multinomial distribution, the Bayesian bootstrap uses Dirichlet distribution. Hence the Bayesian bootstrap can be considered as a smooth version of classical bootstrap. Both converge as N grows large. Dirichlet prior is used for probabilities/proportions, more info [here](http://stats.stackexchange.com/questions/44494/why-is-the-dirichlet-distribution-the-prior-for-the-multinomial-distribution/)
The below shown Bayesian bootstrap has higher dispersion and in case of statistic like `mean`, lower spread is possible by directly using `dirichlet` weights in weighted mean. 

```{r}
set.seed(42)
dat <- c(1,4,9,10)
N <- 1e5

mat1 <- matrix(NA, nrow = N, ncol = length(dat)) # freq
mat2 <- matrix(NA, nrow = N, ncol = length(dat)) # dirichlet
mat3 <- numeric(N)

for(i in 1:N) {
  p.multinomial <- rep(1/length(dat), length(dat)) #default prob of sample function
  mat1[i,] <- sample(dat, length(dat), replace=TRUE, prob=p.multinomial)
  
  p.dirch  <- rdirichlet(1, rep(1, length(dat)))[1,]
  
  #higher spread, but more general (applicable to median etc)
  mat2[i,] <- sample(dat, length(dat), replace=TRUE, prob=p.dirch)
  
  #lower dispersion, can be used only with weighted statistic like mean
  mat3[i] <- weighted.mean(dat, p.dirch)
}
```
<hr>

Using `boot` and `bayesboot` packages:
```{r}
mat4 <- boot(dat, function(d,w) {mean(d[w])}, N)$t[,1] # boot
mat5 <- bayesboot(dat, mean, N)$V1                     # bayesboot
```

<hr>

## Bayesian bootstrap in JAGS

The model for Bayesian bootstrap in JAGS:

```{r, results='hold'}
bayesian.bootstrap.model <- 
  "model {
      p ~ ddirch(rep(1,N))

      for (j in 1:N) {
        pick[j] ~ dcat(p[])
        yboot[j] <- y[pick[j]]
      }

      y.samples <- yboot[]
  }"

jgs = jags.model(file = textConnection(bayesian.bootstrap.model), data = list('y'=dat,'N'=length(dat)), n.adapt = 1000)

update(jgs, 1000)

out = jags.samples(jgs, c('y.samples'), 3*N, 3)

mat6 = out$y.samples %>% as.vector %>% matrix(nrow=N, byrow=TRUE)

```

<hr>
```{r, echo=FALSE}
mu1 <- mean(rowMeans(mat1));sd1 <- sd(rowMeans(mat1))
mu2 <- mean(rowMeans(mat2));sd2 <- sd(rowMeans(mat2))
mu3 <- mean(mat4);sd3 <- sd(mat4)
mu4 <- mean(mat5);sd4 <- sd(mat5)
mu5 <- mean(rowMeans(mat6));sd5 <- sd(rowMeans(mat6))
```

```{r,echo=FALSE}
mats <- list(rowMeans(mat1), rowMeans(mat2), mat3, mat4, mat5, rowMeans(mat6))
titles <- c(paste0("Classic bootstrap, mean = ",round(mean(rowMeans(mat1)),2), " sd = ", round(sd(rowMeans(mat1)),2)),
            paste0("Bayesian bootstrap, mean = ",round(mean(rowMeans(mat2)),2)," sd = ",round(sd(rowMeans(mat2)),2)),
            paste0("Bayesian bootstrap (weighted), mean = ",round(mean(mat3),2), " sd = ", round(sd(mat3),2)),
            paste0("boot package, mean = ",round(mean(mat4),2), " sd = ", round(sd(mat4),2)),
            paste0("bayesboot package, mean = ",round(mean(mat5),2)," sd = ",round(sd(mat5),2)),
            paste0("Bayesian bootstrap (JAGS), mean = ",round(mean(rowMeans(mat6)),2), " sd = ", round(sd(rowMeans(mat6)),2)))
lapply(1:6, function(x) {
  mjs_plot(mats[[x]], width=300, height=300, title=titles[x]) %>%
    mjs_histogram(bar_margin=2) %>% 
    mjs_labs(x_label=sprintf("mat%d", x))
}) -> plots

mjs_grid(plots, nrow=3, ncol=2, widths=c(rep(0.5, 2)))
```

